Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-13 supplementary/ptm_sm.tex
Made updates to formulas
Updated shigella to time series model

The methods described previously in Section \ref{sec:test} can be extrapolated to experiments with complex designs targeting PTMs. This includes experiments with additional sources of variation and unbalanced designs. This is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein. The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$. 
\begin{align*}
H_{0}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &= 0 \\
H_{a}: \Delta = (\mu_{i} - \mu_{i'}) - (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) &\neq 0
\end{align*}

@@

The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = \left[ \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
\]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Dataset 1 : Computer simulation 1 - label free experiment}

@@

\subsubsection{Dataset 2 : Computer simulation 2 - missing values and low replicates}

@@

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
\label{sec:benchmark}

@@

\subsection{Dataset 4 : Human-1mix-TMT - ubiquitination}
\label{sec:ipah}

In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD). \cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. The experimental design can be seen in Table~\ref{table:ipah_design}.

@@

\clearpage
\subsection{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
\label{sec:shigella}

In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed. \cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. The experimental design can be seen in Table~\ref{table:shigella_design}.

@@

\begin{figure}[h!]
\centering
\includegraphics[height=.65\textwidth]{sim_new/shig_venn_diagramm.png}
\caption{The overlap of differential modified peptides between the PTM model with and without global protein level adjustment. Much like in \sfigref{fig:ipha_venn_d}, more PTMs became insignificant after adjustment then became significant. This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 776 peptides became insignificant due to an increase in variance. This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.}
\label{fig:shig_venn_d}
\end{figure}

@@

\subsection{Dataset 6 : Human label free quantification - no global profiling run}
\label{sec:usp30}

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson's Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 overexpression (USP30\_OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.

>>>>

The methods described previously in Section \ref{sec:test} can be extrapolated to experiments with complex designs targeting PTMs. This includes experiments with additional sources of variation and unbalanced designs. This is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein. The null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$. 
\begin{align*}
H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
\end{align*}

@@

The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
\]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Dataset 1 : Computer simulation 1 - Label-free}

@@

\subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}

@@

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
\label{sec:benchmark}

@@

\subsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:ipah}

In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD). \cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. The experiment was modeled as a group comparison, with different replicates measured at each time point. The experimental design can be seen in Table~\ref{table:ipah_design}.

@@

\clearpage
\subsection{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:shigella}

In this study, the correlation between the gene Atg16L1 and killing of Shigella flexneri (S.flexneri) was assessed. \cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). The experiment was modeled as a time series, with the same bioreplicate being measured at each time point. Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. The experimental design can be seen in Table~\ref{table:shigella_design}.

@@

\begin{figure}[h!]
\centering
\includegraphics[height=.65\textwidth]{sim_new/shig_venn_diagramm.png}
\caption{The overlap of differential modified peptides between the PTM model with and without global protein level adjustment. Much like in \sfigref{fig:ipha_venn_d}, more PTMs became insignificant after adjustment then became significant. This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.}
\label{fig:shig_venn_d}
\end{figure}

@@

\subsection{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
\label{sec:usp30}

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson's Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 overexpression (USP30\_OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary. The experiment was modeled as a group comparison.
